---
title: Numerical data
subtitle: ST201 - Data analysis
filters:
  - code-window
---

```{r}
#| label: setup

set.seed(4321)

# Options ----------------------------------------------------------------------
knitr::opts_chunk$set(fig.align = "center")

# Packages ---------------------------------------------------------------------
library(dplyr)
library(ggplot2)
library(tinytable)

# Data -------------------------------------------------------------------------
student_data <- readRDS(here::here("data/student_performance.RDS"))
```


## Recap

Last week, we covered:

- **Categorical** and **numerical** variables.
- **Subtypes** of categorical and numerical variables.
- **Ways to represent** these types of variables.

::: {.callout-important}
## Check in: Was there anything that was confusing or not clear from last week?
:::

## Numerical data {style="font-size: 95%;"}

This week, we're going to talk a little bit more about **numerical data**.

We'll be answering three questions


::: {.fragment .fade-up}

**Where is the data?**

- Mean · Median · Quantiles

:::


::: {.fragment .fade-up}

**How much does the data vary?**

- Range · IQR · Variance · Standard deviation

:::


::: {.fragment .fade-up}

**How are two variables related?**

- Scatterplots · Correlation

:::

## Numerical data {style="font-size: 90%;"}

We'll be working with the same dataset as last week.

<br>

```{r}
#| label: tbl-1
#| tbl-cap: "Student performance data"

# Output data as a nice table --------------------------------------------------
student_data |>
  dplyr::select(
    ID = student_id,
    programme,
    year,
    `Hours of study` = study_hours,
    `Assignments` = assignments_completed,
    satisfaction,
    passed,
    grade
  ) |>
  dplyr::rename_with(function(x) stringr::str_to_sentence(x)) |>
  dplyr::rename(ID = Id) |>
  dplyr::slice_head(n = 5) |>
  tinytable::tt() |>
  tinytable::style_tt(i = 0, bold = TRUE) |>
  tinytable::style_tt(j = 2:7, align = "c")
```

## Numerical data {style="font-size: 90%;"}

We'll be working with the same dataset as last week.

<br>

```{r}
#| label: tbl-2
#| tbl-cap: "Student performance data"

# Output data as a nice table --------------------------------------------------
student_data |>
  dplyr::select(
    ID = student_id,
    programme,
    year,
    `Hours of study` = study_hours,
    `Assignments` = assignments_completed,
    satisfaction,
    passed,
    grade
  ) |>
  dplyr::rename_with(function(x) stringr::str_to_sentence(x)) |>
  dplyr::rename(ID = Id) |>
  dplyr::slice_head(n = 5) |>
  tinytable::tt() |>
  tinytable::style_tt(i = 0, bold = TRUE) |>
  tinytable::style_tt(j = 2:7, align = "c") |>
  tinytable::style_tt(i = 1:5, j = 8, color = "#FF5733", bold = TRUE)
```

## Let's look at student performance

::: {.callout-note icon="false"}
## Suppose we want to describe the grades of students.

What could we say about the grades?
:::

```{r}
#| label: grade-histogram-1

hist(
  student_data$grade,
  xlab = "Student grade",
  main = "Histogram of student grades"
)
```

## The distribution of grades {style="font-size: 80%;"}

A distribution describes **how the values of a variable are spread out**.

- Where values tend to be
- How spread out they are
- Whether there are unusual values
- Whether the distribution is symmetric or skewed

```{r}
#| label: distributions-figure

tibble::tibble(
  normal = rnorm(n = 100, mean = 0, sd = 1),
  gamma = rgamma(n = 100, shape = 0.75),
  uniform = runif(n = 100, min = 0, max = 1)
) |>
  tidyr::pivot_longer(
    cols = dplyr::everything(),
    names_to = "dist",
    values_to = "val"
  ) |>
  dplyr::mutate(
    dist = factor(
      dist,
      levels = c("normal", "gamma", "uniform"),
      labels = c("Normal", "Gamma", "Uniform")
    )
  ) |>
  ggplot(aes(x = val, fill = dist)) +
  geom_histogram(colour = "black", bins = 10, alpha = 0.5) +
  ggokabeito::scale_fill_okabe_ito() +
  labs(x = NULL, y = NULL) +
  facet_wrap(
    ~dist,
    scales = "free",
    labeller = labeller(
      dist = function(x) paste(x, "distribution")
    )
  ) +
  guides(fill = "none") +
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

## The mean

The mean gives us a measure of the **centre** of a distribution.

::: {.fragment}

$$
\bar{x} = \frac{1}{n}\sum^n_{i = 1}x_i
$$

More simply: **add all the values and divide by the number of values.**

:::


::: {.fragment}

$$
60 + 70 + 80 = 210 \\[8pt]
210 / 3 = 70
$$

> If the total grade points were shared equally among students, everyone would
> have a grade of 70.
:::

## Back to the performance dataset

```{r}
#| label: grade-mean
#| echo: true
#| collapse: true
mean(student_data$grade)
```


::: {.fragment .fade-up}

```{r}
#| label: grade-histogram-2

hist(
  student_data$grade,
  xlab = "Student grade",
  main = "Histogram of student grades"
)

abline(v = 64.201)
```

:::

## Mean deception

Imagine we had two groups of students:

- **Group A**: $40, 50, 60, 70, 80, 90, 100$
- **Group B**: $65, 67, 68, 70, 72, 73, 75$


::: {.callout-note icon="false"}
## What can you tell me about these groups?
:::

:::: {.fragment}
::: {.callout-tip icon="false"}
## Both groups have a mean of 70
:::
::::

:::: {.fragment}
::: {.callout-important icon="false"}
## The mean tells us about the centre - but not the spread of data (we'll come back to this)
:::
::::

## The median

The median is the **middle value** when the observations are **ordered
from smallest to largest**.

- Half of the observations are below it.
- Half are above it.

$$
\tilde{x}^{(0.5)} = \begin{cases}
x_{(n + 1)/2}, & \text{if n is odd} \\[8pt]
\frac{1}{2}(x_{n/2} + x_{(n/2)+1}), & \text{if n is even}
\end{cases}
$$


::: {.fragment}
::: {.callout-note icon="false" title="Exercise"}
Calculate the median from the following data:

[11, 10, 26, 10, 9, 10, 17, 6, 8, 13]{.center}

:::

{{< countdown minutes=2 >}}
:::

## The median

<br>

$$
\begin{align*}
&11, 10, 26, 10, 9, 10, 17, 6, 8, 13 \\[8pt]
&6, 8, 9, 10, 10, 10, 11, 13, 17, 26 \\[8pt]
&6, 8, 9, 10, \underbrace{10, 10,} 11, 13, 17, 26 \\[16pt]
&\text{Median} = \frac{10 + 10}{2} = 10
\end{align*}
$$

## The median

```{r}
#| label: median-grade
#| echo: true
#| collapse: true

median(student_data$grade)
```


::: {.fragment .fade-up}

```{r}
#| label: grade-histogram-3

hist(
  student_data$grade,
  xlab = "Student grade",
  main = "Histogram of student grades"
)

abline(v = c(64.201, 65.20))
```

:::

## Comparing the mean and median {style="font-size: 90%;"}

Let's say we take have a group of students with the following data

$$
65, 67, 68, 70, 72, 73, 75
$$


::: {.callout-tip icon="false" title="Summary statistics"}

**Mean:** 70 <br>
**Median:** 70

:::


::: {.fragment}

But what happens when we change one value

$$
65, 67, 68, 70, 72, 73, \mathbf{200}
$$

**What value changes?**

:::


:::: {.fragment}
::: {.callout-tip icon="false" title="Summary statistics"}

**Mean:** 87.86 <br>
**Median:** 70

:::
::::

## Quantiles {style="font-size: 90%;"}

- Quantiles are a generalisation of the idea of the median
- A quantile tells us **where a particular proportion** of the observations fall.


::: {.callout-tip icon="false" title="Understanding quantiles"}

- 25th percentile: 25% of observations are at or below this value.
- 50th percentile: 50% of observations are at or below this value (**median**).
- 75th percentile: 75% of observations are at or below this value.
:::


::: {.fragment}

```{r}
#| label: quantile-example
#| echo: true
#| collapse: true
quantile(student_data$grade)
```

:::

## Quantiles

Let's look at this on the ECDF plot.

```{r}
#| label: grade-ecdf-quantile-1

plot.ecdf(
  student_data$grade,
  xlab = "Student grades",
  main = "ECDF plot of student grades"
)

abline(h = c(0.25))
```

## Quantiles

Let's look at this on the ECDF plot.

```{r}
#| label: grade-ecdf-quantile-2

plot.ecdf(
  student_data$grade,
  xlab = "Student grades",
  main = "ECDF plot of student grades"
)

abline(h = c(0.25, 0.5))
```

## Quantiles

Let's look at this on the ECDF plot.

```{r}
#| label: grade-ecdf-quantile-3

plot.ecdf(
  student_data$grade,
  xlab = "Student grades",
  main = "ECDF plot of student grades"
)

abline(h = c(0.25, 0.5, 0.75))
```

## The spread of the data

- **Group A**: $40, 50, 60, 70, 80, 90, 100$
- **Group B**: $65, 67, 68, 70, 72, 73, 75$


::: {.callout-tip icon="false"}
## Both groups have a mean of 70
:::

::: {.callout-important icon="false"}
## But Group A has much more variability
:::

<br>

:::: {.fragment}
::: {.callout-note icon="false"}
## How do we measure the spread of data?
:::
::::

## The range

The range describes the distance between the smallest and largest values.

$$
\text{Range} = \text{max}(x) - \text{min}(x)
$$

::: {.fragment}

```{r}
#| label: range
#| echo: true
#| collapse: true

range(student_data$grade)
```

:::

## Interquartile range (IQR) {style="font-size: 85%;"}

The IQR describes the spread of the **middle 50%** of observations.

$$
\text{IQR} = Q_3 - Q_1
$$


::: {.fragment .fade-up}

```{r}
#| label: grade-boxplot
boxplot(student_data$grade, ylab = "Student grade")
```

:::

## Variance

We can also measure the spread of data using **variance**

$$
s^2 = \frac{1}{n - 1} \sum^n_{i=1}(x_i - \bar{x})^2
$$

- Take each observation's distance from the mean.
- Square the difference.
- Take an average of all the differences

::::{.fragment .fade-up}
::: {.callout-important icon="false"}
## Standard deviation

We can take the square root of this variance and we're left the **standard deviation**

[**How far away is the data from the mean**]{.center}
:::
::::

## Exercise: Standard deviation

All the below examples have a **mean of 5** - which have the following standard
deviations

[**SD = 3; &nbsp; &nbsp; SD = 8; &nbsp; &nbsp; SD = 15; &nbsp; &nbsp; SD = 30**]{.center}

```{r}
#| label: sd-example
set.seed(1234)

# Generate som data
sd_data <- tibble(
  ex_1 = rnorm(200, 5, 3),
  ex_2 = rnorm(200, 5, 30),
  ex_3 = rnorm(200, 5, 15),
  ex_4 = rnorm(200, 5, 8)
)

# Processing
sd_data <- sd_data |>
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "example",
    values_to = "value"
  ) |>
  mutate(
    example = stringr::str_replace(example, "ex_", "Example "),
    example = factor(example, levels = paste("Example", 1:4))
  ) |>
  group_by(example) |>
  mutate(mean_val = mean(value))

# Plotting ---------------------------------------------------------------------
sd_data |>
  ggplot(aes(x = value, fill = example)) +
  geom_histogram(colour = "black") +
  geom_vline(xintercept = 5, linetype = "dashed", colour = "gray60") +
  labs(x = NULL, y = NULL) +
  facet_wrap(
    ~example,
    labeller = labeller(
      example = function(x) paste(x, "(SD = ?)")
    )
  ) +
  guides(fill = "none") +
  theme_bw()
```

## Scatter plots

In all our previous examples, we've been looking at a **single numerical variable**.
But what if we are interesting in **two numerical variables**

:::: {.fragment}
::: {.callout-note icon="false"}
## Does studying for longer relate to higher grades?
:::

```{r}
#| label: scatter-plot
plot(
  student_data$study_hours,
  student_data$grade,
  xlab = "Hours of study",
  ylab = "Student grade"
)
```

::::

## Wrapping up

- **Means** and **medians** describe the **center** of the data
- **Quantiles** tell you the **position** of data
- **Range, IQR, variance,** and **SD** tell you the **spread** of data
- **Histograms** and **boxplots** can show the **distribution** of data
- **Scatter plots** can show **relationships** between **two** variables
  - [We'll learn more about this later in the semester]{.alert}

::: {.callout-important icon="false"}
## Further reading

Chapter 3 of [Introduction to Statistical and Data Analysis](https://www.academia.edu/42933729/Introduction_to_Statistics_and_Data_Analysis_With_Exercises_Solutions_and_Applications_in_R#outer_page_47) by Christian Heimann & Michael Schomaker Shalabh
:::
