---
title: Schedule
sidebar: false
toc: false
page-layout: full
---

Here's your roadmap for the semester!

- {{< fa person-chalkboard >}} This will be a standard lecture format class
  with some in class participation. **This will usually be the Tuesday class**.
- {{< fa laptop-code >}} This will be active coding workshop where we will
  work through a dataset (both as a class and in groups) to incorporate what we
  previously learned. **This will usually be the Thursday class**.
- {{< fa chalkboard-user >}} This will be an in-person tutorial where you will
  go through a previously assigned practice worksheet with a tutor.
- {{< fa pen-ruler >}} This will be a graded Moodle quiz that will take up a
  part of your continuous assessment.

```{r}
#| label: set-up
#| include: false

# Functions --------------------------------------------------------------------
source(here::here("R/build_schedule_table.R"))

# Packages ---------------------------------------------------------------------
library(dplyr)
library(tinytable)

# Class schedule
schedule <- readxl::read_xlsx(
  here::here("schedule/excel/schedule.xlsx")
) |>
  build_schedule_table()
```

```{r}
#| label: show-table
#| echo: false
#| results: asis
schedule |>
  knitr::kable(
    format = "html",
    escape = FALSE,
    col.names = c(
      "Date",
      "Topic",
      "Slides",
      "Coding",
      "Tutorial",
      "Assignment"
    ),
    table.attr = 'class="schedule-table"'
  ) |>
  kableExtra::kable_styling() |>
  kableExtra::column_spec(
    1,
    width = "20%",
    extra_css = "padding-right: 20px;"
  ) |>
  kableExtra::column_spec(2, width = "48%") |>
  kableExtra::column_spec(3:6, width = "8%")

glue::glue(
  '<i class="schedule-icon schedule-icon--inactive fa-solid fa-laptop-code fa-lg"></i>'
)
```
