---
title: Numerical data
---

```{r}
#| label: setup
#| include: false

purrr::map(list.files(here::here("R"), full.names = TRUE), source)
```

## Learning objectives

- Understand how to summarise numerical data
- Means, medians, ranges, variability, and standard deviations
- Learn how to interpret box plots

## Readings

### Suggested readings

- {{< fa book >}} [Chapter 3](https://www.academia.edu/42933729/Introduction_to_Statistics_and_Data_Analysis_With_Exercises_Solutions_and_Applications_in_R#outer_page_47) of Introduction to Statistics and Data Analysis
- {{< fa link >}} [Descriptive statistics in R](https://statsandr.com/blog/descriptive-statistics-in-r/)
- {{< fa brands youtube >}} [Beginner's Guide to R : Easy Ways to do basic data analysis](https://www.youtube.com/watch?v=cVRkVhQeoKM)

## Slides

::: {.callout-caution}
## Slides will be available on Tuesday 29th September
:::

```{r}
#| label: add-slides
#| echo: false
#| eval: false
#| results: asis

add_slides(slide_id = "/slides/output/slides_2.html")
```
