---
title: The Foundations of Data
page-layout: full
---

```{r}
#| label: setup
#| include: false

purrr::map(list.files(here::here("R"), full.names = TRUE), source)
```

## Learning objectives

- Understand the structure of data
- Understand the different types of data
- Know how to visualise different types of data

## Readings

### Suggested readings

- {{< fa book >}} [Chapter 1](https://www.academia.edu/42933729/Introduction_to_Statistics_and_Data_Analysis_With_Exercises_Solutions_and_Applications_in_R#outer_page_15) and [Chapter 2](https://www.academia.edu/42933729/Introduction_to_Statistics_and_Data_Analysis_With_Exercises_Solutions_and_Applications_in_R#outer_page_28) of Introduction to Statistics and Data Analysis
- {{< fa book >}} [Chapters 1 and 2](https://leanpub.com/os) of OpenIntro
Statistics. This is a slightly more advanced reading, but
*I would highly recommend*!

## Slides

::: {.callout-caution}
## Available on Tuesday 22nd September
:::

```{r}
#| label: add-slides
#| echo: false
#| eval: false
#| results: asis

add_slides(slide_id = "/slides/output/slides_1.html")
```
