---
title: Categorical data analysis (II)
---

```{r}
#| label: setup
#| include: false

purrr::map(list.files(here::here("R"), full.names = TRUE), source)
```

## Learning objectives

- Understand how to further analyse a significant $\chi^2$ result
- Learn about Pearson residuals and Cramér’s V
- Understand what to do when our assumptions are violated

## Readings

### Suggested readings

- {{< fa book >}} [Chapter 4](https://www.academia.edu/42933729/Introduction_to_Statistics_and_Data_Analysis_With_Exercises_Solutions_and_Applications_in_R#outer_page_77) of Introduction to Statistics and Data Analysis
- {{< fa book >}} [Chapter 12](https://learningstatisticswithr.com/12-chisquare.html) of Learning Statistics with R
- {{< fa brands youtube >}} [Statistical Tests for Categorical Data using R](https://www.youtube.com/watch?v=BV9iaJceoHo)

## Slides

::: {.callout-caution}
## Slides will be available on Tuesday 13th October
:::

```{r}
#| label: add-slides
#| echo: false
#| eval: false
#| results: asis

add_slides(slide_id = "/slides/output/slides_4.html")
```
