---
title: Practice Questions Sheet 1
subtitle: ST201 - Data Analysis
author: Cormac Monaghan
description: |
  This tutorial introduces some **basic ways of describing and visualising data**
  in R. Please **attempt all questions** before your tutorial You should work
  through the questions in a new R script.

  For questions asking you to interpret a result, **aim for one or two clear
  sentences**. You should use the context of the pizza delivery data when
  explaining your answers.
format:
  st201_assignment-html: default
  st201_assignment-typst: default
---

```{r}
#| label: setup
#| include: false

# Computations & figures -------------------------------------------------------
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,

  fig.align = "center",
  fig.retina = 3,
  fig.width = 6,
  fig.height = 4,
  out.width = "100%",
  collapse = TRUE
)

# Packages ---------------------------------------------------------------------
library(tinytable)
library(ggplot2)

# The data ---------------------------------------------------------------------
pizza <- readRDS(here::here("assignments/data/pizza_del.rds"))
```

:::: {.content-visible when-format="html"}
::: {.callout-important}
## Please complete this assignment before the tutorial on the week of 12th October
:::

::: {.callout-tip title="Getting help"}
If you are unsure what a function does, remember that you can use R's help system:

`?mean`

You can also use the Help pane in RStudio.
:::
::::

::: {.content-visible when-format="typst"}
```{=typst}

#block(width: 100%, stroke: rgb("#667477"), inset: 1.5em, outset: 0.25em)[
  #text(size: 10pt, weight: "bold")[
    Please complete this assignment before the tutorial on the week of 12th October
]

#v(1em)

_If you cannot remember what a certain function does or a certain function's argument_

#v(1em)

If you are unsure what a function does, remember that you can use R's help system:

#block[
#Skylighting(([#CommentTok("# What does the mean function do?");],
[#FunctionTok("?mean()")]));
]

You can also use the Help pane in RStudio.
]
```
:::

\newpage

# Pizza delivery data

We will use a dataset containing information about **1,301 pizza deliveries**.

The dataset contains **four variables**:

```{r}
#| label: tbl-pizza
#| tbl-cap: "Overview of the pizza dataset"
#| echo: false

pizza_overview <- dplyr::tribble(
  ~Variable      , ~Description                                                         ,
  "Day"          , "Day of the week on which the delivery was made"                     ,
  "Distance"     , "Distance between the pizza restaurant and delivery location, in km" ,
  "Order amount" , "Value of the pizza order"                                           ,
  "Tip"          , "Tip left by the customer"
)

# Output as a table
pizza_overview |>
  tt() |>
  style_tt(i = 0, bold = TRUE) |>
  style_tt(j = 1:2, align = "l")
```

The dataset is available on the RStudio server in the ST201 data folder.

You can load it by opening a new R script and running the following code:

```{r}
#| label: load-pizza
#| eval: false

# Load the pizza dataset
pizza <- readRDS("Sharedfiles/ST201/data/pizza_del.rds")
```

Once you have loaded the data, you can look at the first few observations using:

```{r}
#| label: examine-pizza
#| eval: false

# Examine the first 6 rows of the pizza dataset
head(pizza)
```

## Exercise 1

- How many **observations** are in the dataset?
- How many **variables** are in the dataset?
- Which variables are **categorical** and which are **numerical**?

\newpage

# Individual observations

Let's start by looking at the `distance` variable.

Copy and run the below code

```{r}
#| label: pizza-distance
#| eval: false

pizza$distance[1:10]
```

This displays the first 10 observations of distance.

::: {.callout-tip collapse="true"}
## Remember that R uses square brackets `[]` to select observations.

For example

```{r}
#| label: pizza-distance-1
#| eval: false

# The first observation
pizza$distance[1]
```

selects the **first observation**, while

```{r}
#| label: pizza-distance-10
#| eval: false

# The tenth observation
pizza$distance[10]
```

selects the **tenth observation**.
:::

## Exercise 2

- What is the value of the **275th observation** of `distance`?
- What does this **value represent** in the context of the dataset?
- What is the distance for the **1,000th observation**?

\newpage

# Measuring the centre of the data

Suppose the owner of the pizza restaurant wants to know the typical
delivery distance.

One way to describe the centre of a numerical variable is using the mean.

In @fig-pizza-distance-mean, we can see the full distribution of the pizza
delivery distance variable.

```{r}
#| label: fig-pizza-distance-mean
#| fig-cap: "Histogram showing the distribution of pizza delivery distance (km)"
#| echo: false

# Little plot of our distance variable
pizza |>
  ggplot(aes(x = distance)) +
  geom_histogram(fill = "#6F7F6C", colour = "black", alpha = 0.7) +
  scale_x_continuous(label = scales::label_number(suffix = "km")) +
  labs(x = "Pizza delivery distance (km)", y = "Count") +
  theme_bw()
```

To start off, copy and run the below code

```{r}
#| label: mean-distance-pizza
#| eval: false

mean(pizza$distance)
```

## Exercise 3

- What is the **mean** delivery distance?
- Modify the above code and calculate the **median** delivery distance?
- Write one or two short sentences comparing the mean and median and then in the
  context of the data, compare the two values (are they similar, or
  noticeably different)

\newpage

# Quantiles

The mean and median tell us something about the centre of the data.

We can also use **quantiles** to describe where observations fall within
the distribution.

Run the below code

```{r}
#| label: pizza-quantiles
#| eval: false

quantile(pizza$ditance, )
```

:::: {.content-visible when-format="html"}
::: {.callout-warning}
## Aha!! You should get an warning
:::
::::

::: {.content-visible when-format="typst"}

```{=typst}
#align(center)[
  #text(size: 10pt, weight: "bold")[Aha!! You should get an warning
  ]]
```
:::

## Exercise 4

- Modify the above code to correctly calculate the 25th percentile
- Following this, calculate the 75th percentile
- Provide a short interpretation of these two quantiles, in the context of the
  pizza deliveries

\newpage

# Measuring variance

So far, we've been exploring the centre of `distance`. Let's zoom out a little
and look at the spread of our variable

## Exercise 5

- Calculate the **variance** of the `distance` variable
- Now, calculate the **standard deviation** of the `distance` variable
- In general terms, what is the difference between the term *variance* and the
  term *standard deviation*
- What is the **range** of the `distance` variable

\newpage

# Visualising data

@fig-pizza-tip-boxplot is a box plot of the pizza `tips` variable

```{r}
#| label: fig-pizza-tip-boxplot
#| fig-cap: "Boxplot of pizza tips"
#| echo: false

boxplot(pizza$tip, ylab = "Tip amount")
```

## Exercise 6

- Write `R` code to produce the above plot (remember to add in the **y label**)
- Provide a short overview of what the boxplot is telling us (what is the
  median of the `tip` variable, what is the range, are there any outliers, et.)
- Modify the `R` code you just wrote to produce a histogram of the `tip`
  variable

\newpage

# Comparing delivery days

So far, we have been examining the data across all deliveries.

However, we can actually split this data up a little and look at each separate
day (with the use of the `day` variable).

Suppose the restaurant owner wants to know whether customers spend different
amounts on different days of the week.

Run the below code

```{r}
#| label: pizza-saturday
#| eval: false

# Create a dataset only with pizza deliveries on a Saturday
sat_pizza <- pizza[pizza$day == "Saturday", ]

head(sat_del)
```

::: {.callout-tip}
## This creates a new dataset containing Saturday deliveries only.
:::

## Exercise 7

- Using the new dataset, calculate the mean `order_amount` for Saturday
  deliveries.
- Create a dataset containing **Monday** deliveries only.
- Calculate the mean `order_amount` for Monday deliveries.
- Which day has the larger mean order amount?

\newpage

::: {.callout-note collapse="true" title="Bonus question"}

Creating a bunch of new datasets with different delivery days can be quite
inefficient and tedious.

*Imagine doing it 7 times and running seven different analyses!*

Instead, a neat thing we can do is compare all the deliveries at once

Copy and paste the below code which outputs @fig-pizza-boxplot-grouped

```{r}
#| label: fig-pizza-boxplot-grouped
#| fig-cap: "Overview of pizza tips across all delivery days"

boxplot(tip ~ day, data = pizza, ylab = "Tip amount")
```

The `~` tells `R` to "*Plot tips separately for each value of day*".

- Which day appears to have the **lowest variation in tips**?
- Are there any potential outliers?

:::
