---
title: Welcome to ST201
page-layout: full
---

```{r}
#| label: setup
#| include: false

# Functions --------------------------------------------------------------------
purrr::map(list.files(here::here("R"), full.names = TRUE), source)

# Class timetable --------------------------------------------------------------
timetable <- tibble::tribble(
  ~Time                   , ~Location      , ~`Class Details`           ,
  "Tuesday (3pm - 4pm)"   , "Physics Hall" , "Standard format (slides)" ,
  "Thursday (12pm - 1pm)" , "Physics Hall" , "Practical class (R code)"
)
```

## Module overview

Welcome to second year and welcome to ST201!

We will be covering lots of things data related. This will range from the different types of data, how to summarise and visualise data, perform hypothesis tests, and interpret simple
linear regression models

### Learning outcomes

On successful completion of the module, you should be able to

- Calculate probabilities in R.
- Analyse one and two samples using standard parametric methods.
- Display data using graphical methods.
- Fit, interpret and predict from simple linear regression models.

### Assessment

This module is 100% continuous assessment which is broken down into three parts

- 4 assignments in the form of a **Moodle quiz** (worth 15%)
- 1 mid-term exam, which will be held on **November 10th @3pm** (worth 10%)
- 1 final exam, which will be held on **January 7th @10am** (worth 75%)

::: {.callout-important title="Details for Erasmus students"}
If you are an Erasmus student here for **Semester 1 only**
(September - December), you will not be here for the final exam. With that in
mind, your assignments and mid-term will be worth more marks.

If you are here for the full academic year (semester 1 **AND** semester 2), then
your grades will follow the same breakdown as listed above.
:::

### Timetable

```{r}
#| label: timetable
#| echo: false
#| results: asis
timetable |>
  kableExtra::kbl(
    escape = FALSE,
    align = "llc",
    table.attr = 'class="schedule-table"'
  ) |>
  kableExtra::kable_styling() |>
  kableExtra::column_spec(
    1,
    width = "40%",
    extra_css = "padding-right: 20px;"
  ) |>
  kableExtra::column_spec(2:3, width = "30%")
```

### Tutorials

- There will be 4 tutorials for this module, **one every two weeks**.
- These tutorials will start the **week of the 12th October (week 4)**.
- A sign up sheet will be available for Moodle to pick a time slot. To allow
  adequate time for the department to organise everything, **please ensure you
  fill out this sheet before 30th September**

## Slides

All the slides for this module are created using [Quarto](https://quarto.org/).
In fact, this entire module website is built using Quarto. Personally, I love
Quarto. I do just about everything with it! If you every get a chance, do check
it out (ok ... mini rant over!)

---

There are many **little hidden things** you can do with these slides.

You can click in the slides and navigate through them with <kbd>←</kbd> and
<kbd>→</kbd>.

If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while viewing the
slides you can see a list of slide-specific commands (such as like <kbd>f</kbd> for
fullscreen mode or additionally <kbd>s</kbd> for presenter mode if you want
to see my notes).

Finally, if you want to make a PDF of these slides you can press <kbd>e</kbd>,
followed by <kbd>ctrl</kbd> + <kbd>p</kbd>.

```{r}
#| label: add-slides
#| echo: false
#| eval: true
#| results: asis

add_slides(slide_id = "/slides/output/slides_0.html")
```
